
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Non-classical logics provide a flexible means for describing and reasoning about
diverse computational phenomena.  \emph{Intuitionistic logic} provides a
foundation for constructive mathematics and is the basis of many proof
assistants which are used to prove program correctness.  \emph{Modal logic}
permits succinct reasoning about possible worlds, which has applications to
authorization and distributed computing.  \emph{Linear logic}, with its
interpretation of hypotheses as resources, is used to study stateful systems,
including protocol verification and concurrency.

Theorem proving in these logics is typically difficult. For example,
intuitionisitic propositional logic is PSPACE-complete, while even fragments of
linear propositional logic are undecidable.  Nevertheless, it is fundamental to
the adoption of such expressive logics that we develop practical theorem proving
tools.

We describe a general method for building efficient theorem provers for
non-classical logics called the \emph{polarized inverse method}.  The
\emph{generality} of our approach comes from the use of the inverse method, a
well-known proof search strategy for cut-free sequent calculi, and the use of
\emph{constraints} for delaying the results of costly computations.
\emph{Efficiency} comes from the relatively novel proof-theoretic techniques of
\emph{focusing} and \emph{polarization} as well as more traditional techniques
such as term indexing and redundancy elimination.

Throughout the thesis we will demonstrate by example that the inverse method,
augmented with constraints, focusing, and polarization, forms a strong
foundation for efficient theorem proving in non-classical logics.  We give
support to our claims by describing \emph{Imogen}, a theorem prover for
propositional and first-order intuitionistic logic, the first-order
intuitionistic modal logics K, T, K4, S4, and S5, and two substructural logics
(linear logic and ordered logic).  We find in empirical tests that Imogen
compares favorably with existing implementations.  For ordered logic and some
intuitionistic modal logics, Imogen is the first and only implementation.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "thesis"
%%% End:
